\Title

### 2023-11-23 source/tex/paper.tex
Replaced Dx with \Delta x
Previous versions used both Dx and Delta x for the same thing -- the change in coordinates. Now I use just Delta x

% ============================================================================================
\title{\Title}
\author{\Author\\[10pt]%
School of Mathematical Sciences\\%
Monash University, 3800\\%
Australia}
% \date{10-Nov-2008}% 1st version started
% \date{29-Nov-2008}% 1st version finished
% \date{10-Mar-2009}% 1st version to gr-qc
% \date{22-Mar-2009}% minor corrections to section 8.2
% \date{09-Jul-2009}% minor grammatical corrections in response to 1st referee
% \date{12-Aug-2009}% small corrections arising from proofs received from CQG
\date{12-Aug-2009 (v1)\\
      10-Dec-2019 (v2)\\  % major changes to correct errors in 4th and higher order terms,
                          % all code will be available on github
       4-Nov-2022 (v3)\\  % corrected typos on lhs of eqtns 11.19a-d,
       1-Dec-2022 (v4)}   % corrected typo on equation (3.2) in section (3.1)
\reference{Preprint: arXiv:0903.2087\\[5pt]
           Journal: {\it Class.Quantum.Grav.} {\bf 26} 175017 (2009)}

@@

\bgroup
\BegNote{\bf Version 2}
\small%
The first version of this paper was written in 2009. This second version (written in 2019)
differs from the first in a number of important ways. First, the discussion on Cadabra has
been completely rewritten to account for the revised syntax introduced in the latest version
of Cadabra (itself updated in 2016)\footnote{The Cadabra results in this paper are based on
\CdbVersion}. Second, and more importantly, a significant computational error in the first
version (of this paper) has been corrected. Though the algorithms given in the first version
were correct their implementation in Cadabra (by this author) was incorrect. This lead to
errors first in the quartic and higher order terms in the generalised connections and then,
down the line, to computations built upon the generalised connections (e.g., the initial and
boundary value problem and the arc-length equations). The Cadabra codes have been corrected
and thoroughly cross-checked. A number of minor typographical errors have also been
corrected.
% \EndNote
\vskip 2pt

@@

\BegNote{\bf Version 4}
\small%
Equation (3.2) in Section 3.1 had a small error -- the coefficient in front of the $\Gamma \Gamma$
term should be $(n-1)$ rather than $n$. Other instances of equation (3.2) (in section 4 and 7)
were correct. This was a simple typographical error. All of the Cadabra results are unchanged.
\EndNote
\egroup

@@

Given the explicit expressions for the metric and the geodesic that joins the points $P$ and
$Q$, the length of that geodesic can be computed by way of the integral
\[
L_{PQ} = \int_P^Q\>\left(g_{ab}(x)\frac{dx^a}{ds}\frac{dx^b}{ds}\right)^{1/2}\>ds
\]
Up to this point the parameter $s$ has been taken to be the proper distance along the
geodesic. However, after careful inspection of the geodesic path \eqref{eqn:SolIVP} it is
clear that any uniform scaling of $s$ is allowed. Thus it is possible to re-scale $s$ so
that $s=0$ at $P$ and $s=1$ at $Q$ (of course, the parameter $s$ no longer measures proper
distance so it may be better to use a different symbol, say $\lambda$). Furthermore, a
standard result states that the integrand is constant along the geodesic and can thus be
evaluated at any point such as $P$. The integral is now trivial to evaluate with the result
\[
L^2_{PQ} = \left.g_{ab}(x)\frac{dx^a}{ds}\frac{dx^b}{ds}\right\vert_{P}
\]
which, using previous results, leads to
%
\begin{dmath*}L^2_{PQ} = \cdb{lsq4} + \BigO{\eps^4}\end{dmath*}
%
A simple calculation shows that this result can also be written in the slightly more
suggestive form
\[
L^2_{PQ} = g_{ab}(\bar x) Dx^a Dx^b + \BigO{\eps^4}
\]
where ${\bar x}^a = (x^a_P+x^a_Q)/2$ is the \emph{coordinate} mid-point of the geodesic.
This does raise the question -- Can higher order estimates for $L^2$ (see equation
\ref{eqn:geodesicLSQ}) be obtained by sampling the metric at suitablly chosen points on
(or near) the geodesic? That question will be left for another occassion.

@@

All of the $\BigO{\eps^6}$ Cadabra programs were not overly demanding on computational
resources, taking between 0.5 second and 12 minutes to run and requiring
between 40 and 360 Mbyte of memory (on a Mac Pro (2013) running macOS 10.14.4).

The Cadabra codes and several support scripts are available from the author's GitHub site
\ \url{https://github.com/leo-brewin/riemann-normal-coords}.

@@

@@

\begin{align}
   x^{a}(s) &= x^{a} + s Dx^{a}
                     + (s-s^2) x^{a}_2
                     + (s-s^3) x^{a}_3
                     + (s-s^4) x^{a}_4
                     + (s-s^5) x^{a}_5
                     + \BigO{s^6,\eps^6}\tag{11.20}\pglabel{eqn:bvp6}\\
   x^{a}_2 &= \nx{2}^{a}_2 + \nx{3}^{a}_2 + \nx{4}^{a}_2 + \nx{5}^{a}_2
            + \BigO{\eps^6}\tag{11.20a}\\
   x^{a}_3 &= \nx{3}^{a}_3 + \nx{4}^{a}_3 + \nx{5}^{a}_3 + \BigO{\eps^6}\tag{11.20b}\\
   x^{a}_4 &= \nx{4}^{a}_4 + \nx{5}^{a}_4 + \BigO{\eps^6}\tag{11.20c}\\
   x^{a}_5 &= \nx{5}^{a}_5 + \BigO{\eps^6}\tag{11.20d}
\end{align}
%
\begin{dgroup*}[spread=5pt]
   \dMath{11.20a.1}{    -3 \nx{2}^{a}_2 = \cdb{bvp622}}{eqn:bvp622}{\hfill}
   \dMath{11.20a.2}{   -24 \nx{3}^{a}_2 = \cdb{bvp623}}{eqn:bvp623}{\hfill}
   \dMath{11.20a.3}{  -720 \nx{4}^{a}_2 = \cdb{bvp624}}{eqn:bvp624}{\hskip 1cm}
   \dMath{11.20a.4}{  -360 \nx{5}^{a}_2 = \cdb{bvp625}}{eqn:bvp625}{\hfill}
%
   \dMath{11.20b.1}{   -12 \nx{3}^{a}_3 = \cdb{bvp633}}{eqn:bvp633}{\hfill}
   \dMath{11.20b.2}{  -720 \nx{4}^{a}_3 = \cdb{bvp634}}{eqn:bvp634}{\hskip 1cm}
   \dMath{11.20b.3}{ -1080 \nx{5}^{a}_3 = \cdb{bvp635}}{eqn:bvp635}{\hskip 1cm}
%
   \dMath{11.20c.1}{  -180 \nx{4}^{a}_4 = \cdb{bvp644}}{eqn:bvp644}{\hfill}
   \dMath{11.20c.2}{ -2160 \nx{5}^{a}_4 = \cdb{bvp645}}{eqn:bvp645}{\hskip 1cm}
%
   \dMath{11.20d.1}{  -360 \nx{5}^{a}_5 = \cdb{bvp655}}{eqn:bvp655}{\hskip 1cm}
\end{dgroup*}

@@

\begin{equation*}
   \pglabel{eqn:geodesicLSQ}
   % LCB: which of these is correct?
   % \left(\Delta s\right)^2 = \nD{0} + \nD{2} + \nD{3} + \nD{4} + \nD{5} + \BigO{\eps^6,Dx^6}
   % \left(\Delta s\right)^2 = \nD{0} + \nD{2} + \nD{3} + \nD{4} + \nD{5} + \BigO{\eps^6,Dx^7}
   \left(\Delta s\right)^2 = \nD{0} + \nD{2} + \nD{3} + \nD{4} + \nD{5}
                           + \BigO{\eps^6}\tag{11.21}
\end{equation*}

@@

All of the Cadabra files used in preparing this paper are available at the GitHub repository
\ \url{https://github.com/leo-brewin/riemann-normal-coords}

>>>>

% ============================================================================================
\title{\Title}
\author{\Author\\[10pt]%
School of Mathematics\\%
Monash University, 3800\\%
Australia}
% \date{10-Nov-2008}% 1st version started
% \date{29-Nov-2008}% 1st version finished
% \date{10-Mar-2009}% 1st version to gr-qc
% \date{22-Mar-2009}% minor corrections to section 8.2
% \date{09-Jul-2009}% minor grammatical corrections in response to 1st referee
% \date{12-Aug-2009}% small corrections arising from proofs received from CQG
\date{12-Aug-2009 (v1)\\
      10-Dec-2019 (v2)\\  % major changes to correct errors in 4th and higher order terms,
                          % all code will be available on github
       4-Nov-2022 (v3)\\  % corrected typos on lhs of eqtns 11.19a-d,
       1-Dec-2022 (v4)\\  % corrected typo on equation (3.2) in section (3.1)
      21-Nov-2023 (v5)}   % reformatted Cadabra ouput to print \Delta x rather than Dx
\reference{Preprint: arXiv:0903.2087\\[5pt]
           Journal: {\it Class.Quantum.Grav.} {\bf 26} 175017 (2009)}

@@

\bgroup
\BegNote{\bf Version 2}
\small%
The first version of this paper was written in 2009. This second version (written in 2019)
differs from the first in a number of important ways. First, the discussion on Cadabra has
been completely rewritten to account for the revised syntax introduced in the latest version
of Cadabra (itself updated in 2016)\footnote{The Cadabra results in this paper are based on
\CdbVersion}. Second, and more importantly, a significant computational error in the first
version (of this paper) has been corrected. Though the algorithms given in the first version
were correct their implementation in Cadabra (by this author) was incorrect. This lead to
errors first in the quartic and higher order terms in the generalised connections and later to computations built upon the generalised connections (e.g., the initial and
boundary value problem and the arc-length equations). The Cadabra codes have been corrected
and thoroughly cross-checked. A number of minor typographical errors have also been
corrected.
% \EndNote
\vskip 2pt

@@

\BegNote{\bf Version 4}
\small%
Equation (3.2) in Section 3.1 had a small error -- the coefficient in front of the $\Gamma \Gamma$
term should be $(n-1)$ rather than $n$. Other instances of equation (3.2) (in section 4 and 7)
were correct. This was a simple typographical error. All of the Cadabra results are unchanged.

% \EndNote
\vskip 2pt

\BegNote{\bf Version 5}
\small%
Versions 2 to 4 used both $Dx^a$ and $\Delta x^a$ to record the change in cooridnates between
two nearby points. This version replaces the $Dx^a$ with $\Delta x^a$.
\EndNote
\egroup

@@

Given the explicit expressions for the metric and the geodesic that joins the points $P$ and
$Q$, the length of that geodesic can be computed by way of the integral
\[
L_{PQ} = \int_P^Q\>\left(g_{ab}(x)\frac{dx^a}{ds}\frac{dx^b}{ds}\right)^{1/2}\>ds
\]
Up to this point the parameter $s$ has been taken to be the proper distance along the
geodesic. However, after careful inspection of the geodesic path \eqref{eqn:SolIVP} it is
clear that any uniform scaling of $s$ is allowed. Thus it is possible to re-scale $s$ so
that $s=0$ at $P$ and $s=1$ at $Q$ (of course, the parameter $s$ no longer measures proper
distance so it may be better to use a different symbol, say $\lambda$). Furthermore, a
standard result states that the integrand is constant along the geodesic and can thus be
evaluated at any point such as $P$. The integral is now trivial to evaluate with the result
\[
L^2_{PQ} = \left.g_{ab}(x)\frac{dx^a}{ds}\frac{dx^b}{ds}\right\vert_{P}
\]
which, using previous results, leads to
%
\begin{dmath*}L^2_{PQ} = \cdb{lsq4} + \BigO{\eps^4}\end{dmath*}
%
A simple calculation shows that this result can also be written in the slightly more
suggestive form
\[
L^2_{PQ} = g_{ab}(\bar x) \Dx^a \Dx^b + \BigO{\eps^4}
\]
where ${\bar x}^a = (x^a_P+x^a_Q)/2$ is the \emph{coordinate} mid-point of the geodesic.
This does raise the question -- Can higher order estimates for $L^2$ (see equation
\ref{eqn:geodesicLSQ}) be obtained by sampling the metric at suitablly chosen points on
(or near) the geodesic? That question will be left for another occassion.

@@

All of the $\BigO{\eps^6}$ Cadabra programs were not overly demanding on computational
resources, taking between 0.5 second and 12 minutes to run and requiring
between 40 and 360 Mbyte of memory (on a Mac Studio M1 running macOS 14.1.1).

@@

Note also that in the following equations, $x^a$ denotes the Riemann normal coordinates of
a typical point, $P$, while those of a nearby point $Q$ are denoted by $x^a + \Delta x^a$.
The single exception to this rule occurs in equations (11.18) that displays a coordinate
transformation from generic coordinates, $x^a$, to (non-orthogonal) Riemann normal
coordinates, $y^a$.

@@

\begin{align}
   x^{a}(s) &= x^{a} + s \Dx^{a}
                     + (s-s^2) x^{a}_2
                     + (s-s^3) x^{a}_3
                     + (s-s^4) x^{a}_4
                     + (s-s^5) x^{a}_5
                     + \BigO{s^6,\eps^6}\tag{11.20}\pglabel{eqn:bvp6}\\
   x^{a}_2 &= \nx{2}^{a}_2 + \nx{3}^{a}_2 + \nx{4}^{a}_2 + \nx{5}^{a}_2
            + \BigO{\eps^6}\tag{11.20a}\\
   x^{a}_3 &= \nx{3}^{a}_3 + \nx{4}^{a}_3 + \nx{5}^{a}_3 + \BigO{\eps^6}\tag{11.20b}\\
   x^{a}_4 &= \nx{4}^{a}_4 + \nx{5}^{a}_4 + \BigO{\eps^6}\tag{11.20c}\\
   x^{a}_5 &= \nx{5}^{a}_5 + \BigO{\eps^6}\tag{11.20d}
\end{align}
%
\begin{dgroup*}[spread=5pt]
   \dMath{11.20a.1}{    -3 \nx{2}^{a}_2 = \cdb{bvp622}}{eqn:bvp622}{\hfill}
   \dMath{11.20a.2}{   -24 \nx{3}^{a}_2 = \cdb{bvp623}}{eqn:bvp623}{\hfill}
   \dMath{11.20a.3}{  -720 \nx{4}^{a}_2 = \cdb{bvp624}}{eqn:bvp624}{\hskip 1cm}
   \dMath{11.20a.4}{  -360 \nx{5}^{a}_2 = \cdb{bvp625}}{eqn:bvp625}{\hfill}
%
   \dMath{11.20b.1}{   -12 \nx{3}^{a}_3 = \cdb{bvp633}}{eqn:bvp633}{\hfill}
   \dMath{11.20b.2}{  -720 \nx{4}^{a}_3 = \cdb{bvp634}}{eqn:bvp634}{\hskip 1cm}
   \dMath{11.20b.3}{ -1080 \nx{5}^{a}_3 = \cdb{bvp635}}{eqn:bvp635}{\hskip 1cm}
%
   \dMath{11.20c.1}{  -180 \nx{4}^{a}_4 = \cdb{bvp644}}{eqn:bvp644}{\hfill}
   \dMath{11.20c.2}{ -2160 \nx{5}^{a}_4 = \cdb{bvp645}}{eqn:bvp645}{\hskip 1cm}
%
   \dMath{11.20d.1}{  -360 \nx{5}^{a}_5 = \cdb{bvp655}}{eqn:bvp655}{\hskip 1cm}
\end{dgroup*}

@@

\begin{equation*}
   \pglabel{eqn:geodesicLSQ}
   L^2_{PQ} = \nD{0} + \nD{2} + \nD{3} + \nD{4} + \nD{5}
            + \BigO{\eps^6}\tag{11.21}
\end{equation*}

@@

This paper along with all of the Cadabra files and several support scripts are available on
the GitHub repository \ \url{https://github.com/leo-brewin/riemann-normal-coords}
